\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-28 paper.qmd
Initial work on ading best tSNE and UMAP layouts

hex_grid_with_counts_s_curve2 <- full_join(
  hex_grid2, 
  df_bin_centroids2 |> select(hexID, std_counts), 
  by = c("hex_poly_id" = "hexID"))

@@

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit
#| fig-cap: "Best fit"
#| fig-width: 10
#| fig-height: 10

@@

```{r}
#| label: read-two-curvilinear-clust-nldr
# Read a variety of different NLDR representations of two_curvilinear_clust
# and plot them on same aspect ratio
tsne_two_curvilinear_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_tsne_perplexity_30.rds")

nldr_two_curvilinear_clust1 <- tsne_two_curvilinear_clust |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("a", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

umap_two_curvilinear_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_umap_n-neigbors_15_min-dist_0.1.rds")

nldr_two_curvilinear_clust2 <- umap_two_curvilinear_clust |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour='#a65628') +
  interior_annotation("b", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

phate_two_curvilinear_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_phate_knn_5.rds")

nldr_two_curvilinear_clust3 <- phate_two_curvilinear_clust |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("c") +
  theme(aspect.ratio = 1)

trimap_two_curvilinear_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_trimap_n-inliers_12_n-outliers_4_n-random_3.rds")

nldr_two_curvilinear_clust4 <- trimap_two_curvilinear_clust |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour='#4daf4a') +
  interior_annotation("d") +
  theme(aspect.ratio = 1)

pacmap_two_curvilinear_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_pacmap_n-neighbors_10_init_random_MN-ratio_0.5_FP-ratio_2.rds")

nldr_two_curvilinear_clust5 <- pacmap_two_curvilinear_clust |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("e") +
  theme(aspect.ratio = 1)
```

```{r}
#| label: combine-data-two_curvilinear_clust

error_two_curvilinear_clust_umap <- read_rds("data/two_curvy_clust/error_two_curvy_clust_umap.rds")
error_two_curvilinear_clust_tsne <- read_rds("data/two_curvy_clust/error_two_curvy_clust_tsne.rds")
error_two_curvilinear_clust_phate <- read_rds("data/two_curvy_clust/error_two_curvy_clust_phate.rds")
error_two_curvilinear_clust_trimap <- read_rds("data/two_curvy_clust/error_two_curvy_clust_trimap.rds")
error_two_curvilinear_clust_pacmap <- read_rds("data/two_curvy_clust/error_two_curvy_clust_pacmap.rds")

error_two_curvilinear_clust <- bind_rows(error_two_curvilinear_clust_umap, 
                         error_two_curvilinear_clust_tsne,
                         error_two_curvilinear_clust_phate,
                         error_two_curvilinear_clust_trimap,
                         error_two_curvilinear_clust_pacmap)

error_two_curvilinear_clust <- error_two_curvilinear_clust |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  group_by(method, a1) |>
  filter(MSE == min(MSE)) |>
  ungroup()
```

```{r}
#| label: error-comp-two_curvilinear_clust

error_plot_two_curvilinear_clust <- ggplot(error_two_curvilinear_clust, 
                           aes(x = a1, 
                               y = MSE, 
                               colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  # geom_vline(xintercept = 15, linetype="solid", 
  #            color = "black", linewidth=0.8, alpha = 0.5) +
  scale_x_continuous(breaks = sort(unique(error_two_curvilinear_clust$a1))[append(seq(1, length(unique(error_two_curvilinear_clust$a1)), by = 5), 24)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a', "#ff7f00",'#a65628')) +
  scale_y_log10() +
  ylab("MSE") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

@@

```{r}
#| fig-cap: "Assessing which of the 5 NLDR layouts on the two_curvilinear_clust digit 1 data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-e). Layout c is universally poor. Layouts a that show two close clusters are universally suboptimal. Layout a is the best choice."
#| label: fig-two_curvilinear_clust-mse
#| fig-pos: H
#| out-height: 100%

free(error_plot_two_curvilinear_clust) + wrap_plots(nldr_two_curvilinear_clust1, nldr_two_curvilinear_clust2, nldr_two_curvilinear_clust3, 
                                    nldr_two_curvilinear_clust4, nldr_two_curvilinear_clust5, ncol = 2)
```

>>>>

hex_grid_with_counts_two_curvy2 <- full_join(
  hex_grid2, 
  df_bin_centroids2 |> select(hexID, std_counts), 
  by = c("hex_poly_id" = "hexID"))

@@

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "Best fit"
#| fig-width: 10
#| fig-height: 10

@@

```{r}
#| label: read-two-curvy-clust-nldr
# Read a variety of different NLDR representations of two_curvy_clust
# and plot them on same aspect ratio
tsne_two_curvy_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_tsne_perplexity_30.rds")

nldr_two_curvy_clust1 <- tsne_two_curvy_clust |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("a", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

umap_two_curvy_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_umap_n-neigbors_15_min-dist_0.1.rds")

nldr_two_curvy_clust2 <- umap_two_curvy_clust |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour='#a65628') +
  interior_annotation("b", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

phate_two_curvy_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_phate_knn_5.rds")

nldr_two_curvy_clust3 <- phate_two_curvy_clust |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("c") +
  theme(aspect.ratio = 1)

trimap_two_curvy_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_trimap_n-inliers_12_n-outliers_4_n-random_3.rds")

nldr_two_curvy_clust4 <- trimap_two_curvy_clust |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour='#4daf4a') +
  interior_annotation("d") +
  theme(aspect.ratio = 1)

pacmap_two_curvy_clust <- read_rds("data/two_curvy_clust/two_curvy_clust_pacmap_n-neighbors_10_init_random_MN-ratio_0.5_FP-ratio_2.rds")

nldr_two_curvy_clust5 <- pacmap_two_curvy_clust |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("e") +
  theme(aspect.ratio = 1)
```

```{r}
#| label: combine-data-two_curvy_clust

error_two_curvy_clust_umap <- read_rds("data/two_curvy_clust/error_two_curvy_clust_umap.rds")
error_two_curvy_clust_tsne <- read_rds("data/two_curvy_clust/error_two_curvy_clust_tsne.rds")
error_two_curvy_clust_phate <- read_rds("data/two_curvy_clust/error_two_curvy_clust_phate.rds")
error_two_curvy_clust_trimap <- read_rds("data/two_curvy_clust/error_two_curvy_clust_trimap.rds")
error_two_curvy_clust_pacmap <- read_rds("data/two_curvy_clust/error_two_curvy_clust_pacmap.rds")

error_two_curvy_clust <- bind_rows(error_two_curvy_clust_umap, 
                         error_two_curvy_clust_tsne,
                         error_two_curvy_clust_phate,
                         error_two_curvy_clust_trimap,
                         error_two_curvy_clust_pacmap)

error_two_curvy_clust <- error_two_curvy_clust |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  group_by(method, a1) |>
  filter(MSE == min(MSE)) |>
  ungroup()
```

```{r}
#| label: error-comp-two_curvy_clust

error_plot_two_curvy_clust <- ggplot(error_two_curvy_clust, 
                           aes(x = a1, 
                               y = MSE, 
                               colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  # geom_vline(xintercept = 15, linetype="solid", 
  #            color = "black", linewidth=0.8, alpha = 0.5) +
  scale_x_continuous(breaks = sort(unique(error_two_curvy_clust$a1))[append(seq(1, length(unique(error_two_curvy_clust$a1)), by = 5), 24)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a', "#ff7f00",'#a65628')) +
  scale_y_log10() +
  ylab("MSE") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

@@

```{r}
#| fig-cap: "Assessing which of the 5 NLDR layouts on the two_curvy_clust digit 1 data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-e). Layout c is universally poor. Layouts a that show two close clusters are universally suboptimal. Layout a is the best choice."
#| label: fig-two_curvy_clust-mse
#| fig-pos: H
#| out-height: 100%

free(error_plot_two_curvy_clust) + wrap_plots(nldr_two_curvy_clust1, nldr_two_curvy_clust2, nldr_two_curvy_clust3, 
                                    nldr_two_curvy_clust4, nldr_two_curvy_clust5, ncol = 2)
```

```{r}
#| label: best-fit-tsne-two-curvy

two_curvy_tsne_scaled_obj <- gen_scaled_data(
  data = tsne_two_curvy_clust)

tsne_two_curvy_scaled <- two_curvy_tsne_scaled_obj$scaled_nldr |>
  mutate(ID = row_number())
lim1 <- two_curvy_tsne_scaled_obj$lim1
lim2 <- two_curvy_tsne_scaled_obj$lim2
r2_tsne <- diff(lim2)/diff(lim1)

## hexagon binning to have regular hexagons
hb_obj_tsne_two_curvy2 <- hex_binning(
  data = tsne_two_curvy_scaled, 
  bin1 = 10, 
  r2 = r2_tsne)

a1_2_tsne <- calc_bins_y(
  bin1 = 10, 
  r2 = r2_tsne
)$a1

## Data set with all centroids
all_centroids_df4 <- hb_obj_tsne_two_curvy2$centroids

## Generate all coordinates of hexagons
hex_grid4 <- hb_obj_tsne_two_curvy2$hex_poly

## To obtain the standardise counts within hexbins
counts_df4 <- hb_obj_tsne_two_curvy2$std_cts
df_bin_centroids4 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df4, 
  counts_df = counts_df4) |>
  filter(drop_empty == FALSE) |>
  mutate(b1 = "b1 = 11")

hex_grid_with_counts_two_curvy4 <- full_join(
  hex_grid4, 
  df_bin_centroids4 |> select(hexID, std_counts), 
  by = c("hex_poly_id" = "hexID"))

tsne_data_two_curvy2_with_hb_id <- hb_obj_tsne_two_curvy2$data_hb_id
df_all_two_curvy4 <- dplyr::bind_cols(training_data_two_curvy, tsne_data_two_curvy2_with_hb_id)
df_bin_two_curvy4 <- avg_highd_data(data = df_all_two_curvy4, col_start = "x")

df_b_two_curvy4 <- df_bin_two_curvy4 |>
  dplyr::filter(hb_id %in% df_bin_centroids4$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_two_curvy4 <- df_b_two_curvy4[match(df_bin_centroids4$hexID, df_b_two_curvy4$hb_id),] |>
  dplyr::select(-hb_id)

## Triangulate bin centroids
tr1_object_two_curvy4 <- tri_bin_centroids(
  df_bin_centroids4, x = "c_x", y = "c_y")
tr_from_to_df_two_curvy4 <- gen_edges(
  tri_object = tr1_object_two_curvy4)

## Compute 2D distances
distance_two_curvy4 <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_two_curvy4, 
  start_x = "x_from", 
  start_y = "y_from", 
  end_x = "x_to", 
  end_y = "y_to", 
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_two_curvy4 <- find_lg_benchmark(
  distance_edges = distance_two_curvy4, 
  distance_col = "distance")

distance_df_small_edges_two_curvy4 <- distance_two_curvy4 |>
  filter(distance < benchmark_two_curvy4) ## 0.221

tr_from_to_df_two_curvy4 <- inner_join(
  tr_from_to_df_two_curvy4, distance_df_small_edges_two_curvy4,
  by = c("from", "to"))

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy4,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5) +
  coord_equal() +
  interior_annotation("b")

```{r}
#| label: best-fit-tsne-two-curvy-proj

# Apply the scaling
df_model_data_two_curvy <- bind_rows(data_two_curvy, true_model_two_curvy, df_b_two_curvy4)

model_df <- dplyr::left_join(
  distance_df_small_edges_two_curvy4 |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

two_curvy_proj_tsne_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") + #31a354
  geom_point(
    size = 0.5,
    alpha = 0.5,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.126, 0.125)) + 
  ylim(c(-0.126, 0.125)) +
  interior_annotation("c") +
  theme(
    legend.position = "none"
  )

two_curvy_proj_tsne_all_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = true_model_df_proj, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#2b8cbe") +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") +
  geom_point(
    size = 0.5,
    alpha = 0.5,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.126, 0.125)) + 
  ylim(c(-0.126, 0.125)) +
  interior_annotation("d") +
  theme(
    legend.position = "none"
  )
```

```{r}
#| label: best-fit-umap-tsne
#| fig-cap: "Best fit"
#| fig-width: 10
#| fig-height: 20

nldr_two_curvy_original + nldr_two_curvy_original +
  trimesh_two_curvy_removed1_with_data + trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_umap_model +
  two_curvy_proj_tsne_model +
  two_curvy_proj_umap_all_model +
  two_curvy_proj_tsne_all_model +
  plot_layout(ncol = 2)
```

### 2024-09-28 paper.tex
Initial work on ading best tSNE and UMAP layouts

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/best-fit-1.pdf}

@@

\includegraphics[width=1\textwidth,height=1\textheight]{paper_files/figure-pdf/fig-two_curvilinear_clust-mse-1.pdf}

@@

\caption{\label{fig-two_curvilinear_clust-mse}Assessing which of the 5
NLDR layouts on the two\_curvilinear\_clust digit 1 data is the better
representation using MSE for varying binwidth (\(a_1\)). Colour used for
the lines and points in the left plot and in the scatterplots represents
NLDR layout (a-e). Layout c is universally poor. Layouts a that show two
close clusters are universally suboptimal. Layout a is the best choice.}

>>>>

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/best-fit-umap1-1.pdf}

@@

\includegraphics[width=1\textwidth,height=1\textheight]{paper_files/figure-pdf/fig-two_curvy_clust-mse-1.pdf}

@@

\caption{\label{fig-two_curvy_clust-mse}Assessing which of the 5 NLDR
layouts on the two\_curvy\_clust digit 1 data is the better
representation using MSE for varying binwidth (\(a_1\)). Colour used for
the lines and points in the left plot and in the scatterplots represents
NLDR layout (a-e). Layout c is universally poor. Layouts a that show two
close clusters are universally suboptimal. Layout a is the best choice.}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/best-fit-umap-tsne-1.pdf}
